Tests of measurement invariance without subgroups: A~generalization of classical methods

### 2012-04-30 www/paper.Rnw
Revisions: o Addressed reviewers' final comments o Re-ran simulation with n=50 condition o Added two new references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@127 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% TODO 1. discussion of measurement vs structural parameters
%%      2. discussion of longitudinal MI
%%      3. re-run simulations
\title{Tests of measurement invariance without subgroups:
  A~generalization of classical methods}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

@@

In general terms, a set of scales is defined to be measurement
invariant with respect to an auxiliary variable $V$ if:
\begin{equation}
    \label{eq:midef}
      f({\bf X} | T, V, \dots) = f({\bf X} | T, \dots),
\end{equation}
where ${\bf X}$ is a data matrix, $T$ is the latent
variable that the scales purport to measure, and $f$ is the model's
distributional form \cite{Mel89}.  In the parametric,
factor-analytic framework
adopted here, this implies that the measurement parameters are equal
across subgroups of individuals and, thus, do not vary with any
variable $V$ \cite{Mer93}.

@@

\subsection{Local Alternatives}
Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
under specific alternatives 
of parameter instability. In particular, we can assume that the pattern
of deviation $\bm{\theta}(v_i)$ can be described as a constant
parameter plus some non-constant deviation $\bm{g}(i/n)$:
\begin{equation}
    \label{eq:local}
  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
\end{equation}
In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
Equation~\eqref{eq:score} do not have zero expectation but rather
\begin{equation}
    \label{eq:localE}
  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
\end{equation}
In this case, the covariance matrix $\bm{C}(\bm{\theta})$ from the expected
outer product of gradients is
\begin{equation}
    \label{eq:localS}
  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
\end{equation}
which, at $\bm{\theta}_0$, coincides with the expected information matrix.

@@

@@

The proposed tests also 
readily extend to other popular psychometric models.  
For example, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  Under this framework, it is possible to assess whether
paths between 
latent variables are stronger for some individuals than for
others.  Further, in applying the proposed tests to factor mean
parameters (as was done in the example), implications of measurement
invariance violations for structural equation modeling may be better
understood.

The proposed tests also naturally extend to the study of differential
item functioning (DIF) 
in item response models (e.g., \citeNP{KopZei10}, who focused on Rasch
models).   
Traditional DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.
%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
%% studying both  measurement invariance in factor analysis and differential item
%% functioning in item response.
While factor-analytic measurement invariance methods and DIF methods
have developed 
largely independently of one another, the methods can certainly be
treated from a unified perspective (e.g.,
\citeNP{Mcd99,Mil11,StaChe06}).   
The tests proposed here were designed with this perspective in mind.

>>>>

%%      4. deal with color in figures
\title{Tests of measurement invariance without subgroups:
  A~generalization of classical methods}
\twoauthors{Edgar C. Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

@@

In general terms, a set of scales is defined to be measurement
invariant with respect to an auxiliary variable $V$ if:
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f({\bf x}_i | t_i, \dots),
\end{equation}
where $x_i$ is the data vector for individual $i$, $t_i$ is the latent
variable for individual $i$ that the scales purport to measure, and
$f$ is the model's 
distributional form \cite{Mel89}.  
We adopt a parametric,
factor-analytic framework here, so that the above equation being true
for all $V$ implies that the measurement
parameters are equal 
across individuals and, thus, do not vary with any
$V$ \cite{Mer93}.

@@

\subsection{Local Alternatives}
Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
under specific alternatives 
of parameter instability. In particular, we can assume that the pattern
of deviation $\bm{\theta}(v_i)$ can be described as a constant
parameter plus some non-constant deviation $\bm{g}(i/n)$:
\begin{equation}
    \label{eq:local}
  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
\end{equation}
In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
Equation~\eqref{eq:score} do not have zero expectation but rather
\begin{equation}
    \label{eq:localE}
  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
\end{equation}
The covariance matrix $\bm{C}(\bm{\theta})$ from the expected
outer product of gradients is then 
\begin{equation}
    \label{eq:localS}
  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
\end{equation}
which, at $\bm{\theta}_0$, coincides with the expected information matrix.

@@

We included a factor mean in the application in order to
demonstrate the tests' generality.  To deal with 
measurement invariance specifically, however, we may elect to focus on
only the 
measurement parameters within the model (omitting the factor mean).
That is, it would 
not have been notable or surprising if we had observed that
the factor mean lacked measurement invariance with respect to GPA:
this result would have implied that individuals' latent
intelligence fluctuates with GPA.  We speculate that such a result
would often be obtained when $V$ is related to the latent variable of
interest.

@@

The proposed tests also 
readily extend to other psychometric models.  
For example, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  This includes the study of measurement invariance
in second-order growth models and related models for longitudinal
data (e.g., \citeNP{FerBal08,McA09}).
The issue of measurement invariance is important in these models in
order to establish that a ``true'' change has occurred in the 
individuals to which the model has been fitted.  The tests described
here can be used to help establish measurement invariance with respect
to both continuous and categorical auxiliary variables, using only
output from the estimated model of interest.

Finally, the tests can be used for the study of differential
item functioning (DIF) 
in item response models (e.g., \citeNP{KopZei10}, who focused on Rasch
models).   
Traditional DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.
The tests proposed here can detect subgroups automatically, however, offering a unified way of
studying both  measurement invariance in factor analysis and differential item
functioning in item response.
While factor-analytic measurement invariance methods and DIF methods
have developed 
largely independently of one another, the methods can certainly be
treated from a unified perspective (e.g.,
\citeNP{Mcd99,Mil11,StaChe06}).   
The tests proposed here were designed with this perspective in mind.
